---
title: "EDS Week 5 Discussion"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Load packages

```{r}
library(tidyverse)
library(sf)
library(terra)
library(spData)
library(spDataLarge)
```

## Load data

```{r}
# Load raster data representing grain sizes with the three classes clay, silt and sand
grain <- terra::rast(system.file("raster/grain.tif", package = "spData"))

nz_height <- spData::nz_height
```

## Part 1:

1.  Subset points higher than 3100 meters in nz_height (from spData)
2.  Create a template raster with rast(), where the resolution is 3 km x 3 km for the extent of the subset dataset
3.  Count numbers of the highest points in each grid cell
4.  Find the maximum elevation in each grid cell

```{r}
# subset points higher than 3100 m in nz_height 
nz_height_3100 <- nz_height %>% 
  filter(elevation > 3100)


# Create template: define the extent, resolution, and CRS based on nz_height_3100
nz_template <- rast(terra::ext(nz_height_3100), 
                    resolution = 3000, 
                    crs = terra::crs(nz_height_3100))


# Convert vector points to raster data
# Function "length" returns a count of the elevation points per cell
nz_raster <- rasterize(nz_height_3100, nz_template, 
                       field = "elevation", fun = "length")

plot(nz_raster, main = "Number of Elevation Points > 3100 in Each Grid Cell")
plot(st_geometry(nz_height_3100), add = TRUE)


# Find the maximum elevation in each grid cell
nz_max <- rasterize(nz_height_3100, nz_template, 
                    field = "elevation", fun = "max")

plot(nz_max, main = "Maximum Elevation in Each Grid Cell ")
plot(st_geometry(nz_height_3100), add = TRUE)
```

## Part 2:

-   With the previous raster, complete the following:
    -   Aggregate the raster that counts the highest points in New Zealand/Aotearoa
    -   Reduce its geographic resolution by half, such that cells are 6 x 6 km
    -   Plot the result
-   Resample back to the original resolution of 3 km x 3 km

```{r}
# Aggregate the raster that counts the highest points in New Zealand/Aotearoa
nz_raster_agg <- terra::aggregate(nz_raster, fact = 2, 
                                  fun = "sum", na.rm = TRUE)

plot(nz_raster_agg, 
                 main = "Aggregated Raster")


# Resample back to the original resolution of 3 km x 3 km
nz_raster_resample <- terra::resample(nz_raster_agg, nz_template)

plot(nz_raster_resample, 
                      main = "Resampled Raster")


# plot both 
plots <- c(nz_raster, nz_raster_resample)
labs <- c("Original 6 x 6 km", "Resample 6 x 6 km")
plot(plots, main = labs)
```

## Part 3:

- Polygonize `grain` and filter to only keep squares that represent clay

```{r}
# plot the grain rasters
plot(grain, main = "Grain Raster")

# Polygonize grain
grain_poly <- terra::as.polygons(grain) %>% 
  st_as_sf()

plot(grain_poly, main = "Grain Polygons")


# Filter to only keep squares that represent clay
grain_clay <- grain_poly %>% 
  filter(grain == "clay")
```















